### 2025-02-19 paper/paper.tex
Finished text-edition of severe case

In this scenario, the predictor variables are moderately correlated, introducing a controlled level of multicollinearity. This setup challenges the assumptions of Ordinary Least Squares (OLS) regression and allows us to assess how Principal Component Regression (PCR) and Partial Least Squares (PLS) handle collinearity and improve prediction performance.

@@

In this scenario, the dataset contains only 40 observations with 10 predictors, where the predictors exhibit strong multicollinearity. The additional predictors are generated via a linear transformation of a base predictor matrix, leading to extreme correlations. Furthermore, only three predictors contribute to the response variable, while the remaining predictors act as redundant noise. This setup presents an extreme case of multicollinearity, making it a challenging setting for regression models.

@@

PCR and PLS attempt to mitigate this issue by selecting fewer components. However, when using too few components (PCR(1) and PLS(1)), the model oversimplifies the predictor space, leading to high bias. As more components are included, the model performance improves significantly. PLS needs the same number of components to reach a similar level of accuracy as PCR (Table 1).

At their optimal points, both PCR and PLS still exhibit some level of bias, as the component selection inherently introduces regularization. However, this bias is outweighed by the substantial reduction in variance, leading to a much lower overall MSE compared to OLS. This highlights the key advantage of regularization-based methods: while they sacrifice some degree of unbiasedness, they compensate by stabilizing predictions and significantly improving generalization performance.

The stability of coefficient estimates also supports these findings. OLS estimates show extreme variability, confirming its inefficiency in this setting. PCR estimates stabilize as more components are included but still suffer from inefficiencies due to noise retention when too few components are used. PLS maintains stable coefficient estimates even with fewer components due to its response-driven selection. However, when the appropriate number of components is selected, both methods achieve comparable stability.

OLS fails to provide reliable estimates in the presence of severe multicollinearity with low observations due to excessive variance in coefficient estimates. PCR helps reduce multicollinearity effects, but selecting too few components results in high bias, while selecting too many components may introduce noise from redundant predictors. PLS, leveraging its supervised nature, achieves strong predictive performance with fewer components. However, when PCR is given enough components, it reaches a similar level of predictive accuracy, making both methods viable solutions depending on the data and modeling constraints.

At their optimal component selections, PCR and PLS remain slightly biased but achieve significantly lower variance compared to OLS, leading to a much lower overall MSE. This reinforces the effectiveness of regularization-based methods in controlling variance and improving model stability, particularly in high-dimensional, low-sample-size settings.

This dataset highlights the extreme limitations of OLS and demonstrates the importance of feature selection and dimensionality reduction in challenging regression problems. Both PCR and PLS emerge as effective methods for handling highly collinear predictors in small datasets, achieving similar results through different mechanisms—PLS using fewer components due to its supervised nature, while PCR requiring slightly more components to reach the same performance.

@@

The dataset used in this study consists of NIR spectra collected from 80 corn samples, measured using the mp5 NIR spectrometer. Each spectrum contains 700 data points recorded in the wavelength range of 2498–1100 nm at 2 nm intervals. The target variable, moisture content, serves as the dependent variable. This dataset and methodology align closely with the results reported in the referenced paper, where similar approaches were applied to corn data. The high-dimensional nature of the spectral data makes this a challenging regression problem, particularly due to the presence of collinearity among the predictor variables (Figure 4).

>>>>

In second scenario, the predictor variables are moderately correlated, introducing a controlled level of multicollinearity. This setup challenges the assumptions of Ordinary Least Squares (OLS) regression and allows us to assess how Principal Component Regression (PCR) and Partial Least Squares (PLS) handle collinearity and improve prediction performance.

@@

In our last scenario, the dataset contains only 40 observations with 10 predictors, where the predictors exhibit strong multicollinearity. The additional predictors are generated via a linear transformation of a base predictor matrix, leading to extreme correlations. Furthermore, only three predictors contribute to the response variable, while the remaining predictors act as redundant noise. This setup presents an extreme case of multicollinearity, making it a challenging setting for regression models.

@@

PCR and PLS attempt to mitigate this issue by selecting fewer components. However, using too few components (PCR(1) and PLS(1)), the model oversimplifies the predictor space, leading to high bias. As reaching optimal number of components, the model performance improves significantly. It is worth noting that, PLS needs the same number of components to reach a similar level of accuracy as PCR (Table 1).

At their optimal points, both PCR and PLS still exhibit some level of bias, as the component selection inherently introduces regularization. However, this bias is outweighed by the substantial reduction in variance, leading to a much lower overall MSE compared to OLS (Figure 3). This highlights the key advantage of dimension-reduction methods: while they sacrifice some degree of unbiasedness, they compensate by stabilizing predictions and significantly improving generalization performance.

The stability of coefficient estimates also supports these findings. OLS estimates show extreme variability, confirming its inefficiency in this setting. PCR estimates stabilize as more components are included but still suffer from inefficiencies due to noise retention when too few components are used. PLS maintains stable coefficient estimates even with fewer components due to its response-driven selection. However, when the appropriate number of components is selected, both methods achieve comparable stability (Figure 16).

Overall, OLS fails to provide reliable estimates in the presence of severe multicollinearity with low observations due to excessive variance in coefficient estimates. PCR helps reduce multicollinearity effects, but selecting too few components results in high bias, while selecting too many components may introduce noise from redundant predictors. PLS, leveraging its supervised nature, achieves strong predictive performance with fewer components. However, when PCR is given enough components, it reaches a similar level of predictive accuracy, making both methods viable solutions depending on the data and modeling constraints (Figure 14 and Table 2).

At their optimal component selections, PCR and PLS remain slightly biased but achieve significantly lower variance compared to OLS, leading to a much lower overall MSE. This highlighted by the effectiveness of both methods in controlling variance and improving model stability, particularly in high-dimensional, low-sample-size settings (Table 1).

Results of all three scenarios highlight the extreme limitations of OLS and demonstrates the importance of dimensionality reduction in challenging regression problems. Both PCR and PLS emerge as effective methods for handling highly collinear predictors in small datasets, achieving similar results through different mechanisms—PLS using fewer components due to its supervised nature, while PCR requiring slightly more components to reach similar performance.

@@

The dataset used in this study consists of NIR spectra collected from 80 corn samples, measured using the mp5 NIR spectrometer. Each spectrum contains 700 data points recorded in the wavelength range of 2498–1100 nm at 2 nm intervals\footnote{This dataset is available at https://eigenvector.com/resources/data-sets/}. The target variable, moisture content, serves as the dependent variable. This dataset and methodology align closely with the results reported in the referenced paper, where similar approaches were applied to corn data [12]. The high-dimensional nature of the spectral data makes this a challenging regression problem, particularly due to the presence of collinearity among the predictor variables (Figure 4).
